AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology

### 2013-06-02 paper.tex
lista de autores ficou baoa
colocado diagrama sem corte brusco aa direita
retiradas incidencias de primeiras pessoas sing e pl

%
% Final Checklist TODO XXX 
% - Spellcheck
% - check xxx's todo's etc
%
%\input{preamble}
\usepackage{url}
\usepackage{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{color}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
%\usepackage{draftwatermark}

@@

Despite the advantages of GSD, we have noticed how
difficult it is to coordinate and fund free software on a significantly larger scale
than currently practiced, as series of qualitatively new situations arise. Distributed
teams are very heterogeneous containing not only volunteers and very experienced
developers, but also contractors and freelancers from different backgrounds and
cultures. Our observations are founded on the factors suggested by
Carmel~\cite{carmel1999} as main difficulties for GSD: distance, time and
cultural differences. In the case of free or open software projects, all these
factors are involved.

Another problem faced by modern software companies and other
collectives are frequent ineffective meetings, which are seldom
focused on the particular interest of any attendant. The result is that it has
become the norm to participate in too many meetings with the ``laptop
open'', which can be un-productive. Software developers
like to code, to be productive, to have their hands on their project,
to do what they are best at. They dislike to forcibly stop for meetings
or to do other bureaucratic activities such as writing lengthy reports to
justify their funding.\cite{Thompson:Wired:2012}
%\todo{ler mais. cacm, etc}

In this paper we propose the AA methodology and an associated software system
for coordinating distributed team work, tackling the disadvantages
of GSD. Team members take on an egalitarian role, and stay
voluntarily \textit{logged} in the system for part of their time
(e.g.\ 2 hours per day), during which they log a periodical short text
sentence or \emph{microlog} --- similar to a `tweet' from Twitter --- as the
status of their activity. Logging is carried out using a series of
easy UI alternatives: unix shell
commands, native GUI or web page, conventional social network posts, or chat messages to a log bot
listening to IRC, GTalk, G+, and others.  These ``microblog sentences'' are
publicly aggregated and validated by other team members. Through AA, the
community has a
methodology and an associated system to help
implement and validate the activities of a distributed software team. It implicitly
legitimizes financial support for the expansion of the activity of a
distributed development team. The AA methodology is specially useful for
coordinating distributed and decentralized team work, providing effective means
to asynchronously update different team members without the need for synchronous
unproductive meetings.

@@

A brief overview of current work on GSD methodologies related to AA is presented
in Section~\ref{related-work}. In Section~\ref{aa-methodology} the most relevant
characteristics of the AA methodology are outlined. In the same section, we
report an actual use case of AA for coordinating a team of 9 paid developers 
% nao era mais de 9?
during the second half of 2011, as well as a broader use case of AA from 2012 to
the present time.  In Section~\ref{conclusions} we summarize the overall conclusions
and indicate future possibilities for the practical use of AA in other types of
teams of software developers or organizations working on non-software distributed
activities.

@@

Some of the strategies for GSD mentioned on the previous section are based on
complex methodologies and many were built for a specific company or software
center. We propose an alternative methodology based on a simple idea: small
working sessions logged by a computational tool. Figure~\ref{fig:mm} summarizes
this methodology.

\begin{figure}
\begin{center}
   \includegraphics[width=0.8\linewidth,keepaspectratio=true]{figs/aa-mm.png}
\end{center}
   \caption{
   A mindmap of the AA methodology: 1) `developer engagement cycle', \ie, his use
   of AA; 2) `functionality', \ie, main goals of the system; 3) `potentialities',
   \ie, enhancements AA delivers to the developer's context. 
   }
\label{fig:mm}
\end{figure}

@@

Developers optionally record a brief \emph{video screencast} at the end of the
session summarizing what has been done, explaining his goals and challenges
in his own words and showing his most important results. This is very similar to
the video logging system in the movie Avatar, although it is clear from our
July 2011 git repositories and online wiki that we have used this powerful
concept in AA prior to the release of Avatar. Moreover, screencasting differs
from general videologging in that it typically captures actual workflow on the
computer screen. Screencasting, combined with the textual log of the AA Session,
renders the final report more understandable for the individual developer
himself or to other people searching for information about his production.
% TODO: talvez figura com diagraminha explicando uma semana tipica,
% e cada sessao AA com um bloco dividido nos 15 min, e no inicio do bloco tb
% podemos listar no diagrama ``developer reads felow AA logs and lists his
% assigned tickets before he starts''

@@

Peer validation also helps in making decentralized collaboration more cohesive
by encouraging members to be minimally aware of peer activities, even when these
are not immediately useful for accomplishing the task at hand.  We have observed
that decentralized teamwork can get so efficient at actual production that 
the team gets short-sighted in terms of coordination: non-communicating
subteams can get formed in practice if care is not taken, causing a
fragmentation of the collective. Peer validation is one way to help avoid
fragmentation and is an essential mechanism of decentralized team autoregulation.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. We applied the proposed methodology to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what we called Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

We have analyzed data and gained practical experience on
the use of AA to auto-regulate the work of Lab Macambira, a group of free
software developers from Brazil. Since July of 2011 the group has contributed
and created new free and open source software for a vast number of additional applications.

@@

\section*{Acknowledgments}
The authors acknowledge the financial support from Pontao Nos Digitais, and
Ricardo Fabbri acknowledges support from FAPERJ/Brazil 111.852/2012.
We would also like to thank AA: the present research and even this
manuscript was written using AA. The complete log is on-line at
\url{www.pulapirata.com/skills/aa}.

>>>>

%
% Final Checklist TODO XXX 
% - Spellcheck
% - check xxx's todo's etc
%
%\input{preamble}
\usepackage{url}
\usepackage{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{color}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{authblk}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
%\usepackage{draftwatermark}

@@

Despite the advantages of GSD, it is known how
difficult it is to coordinate and fund free software on a significantly larger scale
than currently practiced, as series of qualitatively new situations arise. Distributed
teams are very heterogeneous containing not only volunteers and very experienced
developers, but also contractors and freelancers from different backgrounds and
cultures. Our observations are founded on the factors suggested by
Carmel~\cite{carmel1999} as main difficulties for GSD: distance, time and
cultural differences. In the case of free or open software projects, all these
factors are involved.

Another problem faced by modern software companies and other
collectives are frequent ineffective meetings, which are seldom
focused on the particular interest of any attendant. The result is that it has
become the norm to participate in too many meetings with the ``laptop
open'', which can be un-productive. Software developers
like to code, to be productive, to have their hands on their project,
to do what they are best at. They dislike to forcibly stop for meetings
or to do other bureaucratic activities such as writing lengthy reports to
justify their funding~\cite{Thompson:Wired:2012}.
%\todo{ler mais. cacm, etc}

To address these matters is the purpose of
AA methodology and an associated software system
for coordinating distributed team work, tackling the disadvantages
of GSD. Team members take on an egalitarian role, and stay
voluntarily \textit{logged} in the system for part of their time
(e.g.\ 2 hours per day), during which they log a periodical short text
sentence or \emph{microlog} --- similar to a `tweet' from Twitter --- as the
status of their activity. Logging is carried out using a series of
easy UI alternatives: unix shell
commands, native GUI or Web page, conventional social network posts, or chat messages to a log bot
listening to IRC, GTalk, G+, and others.  These ``microblog sentences'' are
publicly aggregated and validated by other team members. Through AA, the
community has a
methodology and an associated system to help
implement and validate the activities of a distributed software team. It implicitly
legitimizes financial support for the expansion of the activity of a
distributed development team. The AA methodology is specially useful for
coordinating distributed and decentralized team work, providing effective means
to asynchronously update different team members without the need for synchronous
unproductive meetings.

@@

A brief overview of current work on GSD methodologies related to AA is presented
in Section~\ref{related-work}. In Section~\ref{aa-methodology} the most relevant
characteristics of the AA methodology are outlined. In the same section, it
is reported an actual use case of AA for coordinating a team of 9 paid developers 
% nao era mais de 9?
during the second half of 2011, as well as a broader use case of AA from 2012 to
the present time.  In Section~\ref{conclusions}, there are overall conclusions
and indicative of future possibilities for the practical use of AA in other types of
teams of software developers or organizations working on non-software distributed
activities.

@@

Some of the strategies for GSD mentioned on the previous section are based on
complex methodologies and many were built for a specific company or software
center. This section describes an alternative methodology based on a simple idea: small
working sessions logged by a computational tool. Figure~\ref{fig:mm} summarizes
this methodology.

\begin{figure}
\begin{center}
   \includegraphics[width=0.8\linewidth,keepaspectratio=true]{figs/aa-mm2.png}
\end{center}
   \caption{
   A mindmap of the AA methodology: 1) `developer engagement cycle', \ie, his use
   of AA; 2) `functionality', \ie, main goals of the system; 3) `potentialities',
   \ie, enhancements AA delivers to the developer's context. 
   }
\label{fig:mm}
\end{figure}

@@

Developers optionally record a brief \emph{video screencast} at the end of the
session summarizing what has been done, explaining his goals and challenges
in his own words and showing his most important results. This is very similar to
the video logging system in the movie Avatar, although it is clear, from
July 2011 git repositories and online wiki, that we have used this powerful concept in AA independently of any major mass midia incident. Moreover, screencasting differs
from general videologging in that it typically captures actual workflow on the
computer screen. Screencasting, combined with the textual log of the AA Session,
renders the final report more understandable for the individual developer
himself and to other people searching for information about his production.
% TODO: talvez figura com diagraminha explicando uma semana tipica,
% e cada sessao AA com um bloco dividido nos 15 min, e no inicio do bloco tb
% podemos listar no diagrama ``developer reads felow AA logs and lists his
% assigned tickets before he starts''

@@

Peer validation also helps in making decentralized collaboration more cohesive
by encouraging members to be minimally aware of peer activities, even when these
are not immediately useful for accomplishing the task at hand.  Authors have observed
that decentralized teamwork can get so efficient at actual production that 
the team gets short-sighted in terms of coordination: non-communicating
subteams can get formed in practice if care is not taken, causing a
fragmentation of the collective. Peer validation is one way to help avoid
fragmentation and is an essential mechanism of decentralized team autoregulation.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applyed to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

Authors have analyzed data and gained practical experience on
the use of AA to auto-regulate the work of Lab Macambira, a group of free
software developers from Brazil. Since July of 2011 the group has contributed
and created new free and open source software for a vast number of additional applications.

@@

\section*{Acknowledgments}
Authors acknowledge the financial support from Pontao Nos Digitais, and
Ricardo Fabbri acknowledges support from FAPERJ/Brazil 111.852/2012.
Authors also thank AA: the present research and even this
manuscript was written using AA. The complete log is on-line at
\url{www.pulapirata.com/skills/aa}. Finally, authors are also grateful to IFSC/USP, IPRJ/UERJ, IFT/UNESP and all AA users and collaborators, especially those who coded AA hacks for use through shell and bot, and those who coded different Web interfaces in use.
